Advanced Normalization Tools (ANTS)

### 2011-02-09 Documentation/Tex/ants.tex
ENH: updating BTP example
git-svn-id: https://advants.svn.sourceforge.net/svnroot/advants@773 fd2da0a4-d968-48df-9cb5-d8c496287cbf

\subsection{The ANTS Executable}
The ANTS program itself is the central program encapsulating
normalization/registration functionality.  Its main output is an
affine transform file and a deformation field, potentially with
inverse.  Options to ANTS allows the user to navigate the similarity
and transformation options that are available.  ANTS takes multiple
similarity and optimization criteria as options.  The program is
wrapped in \verb ants.sh ~for normalization with ``out of the box''
parameters and in \verb buildtemplateparallel.sh ~for 
computationally distributed optimal template construction.

\subsection{The WarpImageMultiTransform Executable}
The WarpImageMultiTransform program applies ANTS mappings to images.  One may apply an 
arbitrarily long series of transformations to an image through this program.  Thus, it enables one 
to compose a series of affine and deformable mappings -- and/or their inverses -- such that one 
may avoid multiple interpolations of a single image.  This program is used internally in ANTS 
optimization to avoid ever using more than one interpolation.  The use-NN option applies nearest 
neighbor interpolation, otherwise linear interpolation is used.

@@

\subsection{The ImageMath Executable} This is a multi-purpose program that has the following syntax:  ``ImageMath ImageDimension outputfilename   Operation  InputFileName  parameters ``.  Most basic scalar image operations -- and some tensor operations -- may be performed with this program.  Some operations output text files, some output images and some output only to the terminal.   ImageMath allows one to multiply images together (m), to negate images (Neg), to take an image to a power (pow), to test the invertibility of transformations (InvId), to compute the fractional anisotropy of an image (TensorFA) and to perform Markov Random Field segmentation with or without priors (Segment).   Many other operations are available.  Like all other ANTS programs, one may call ImageMath from the command line to see all of its options.   ImageMath is used heavily in ANTS scripts.

\subsection{ANTS Scripts} 
The ANTS/Scripts directory contains (hopefully) user-friendly
wrappings of ANTS tools that enable higher-level error checking and
combinations of basic ANTS functions.  These scripts are called as
\verb sh ~ \verb antsscriptname.sh ~ and provide usage when called
from the command line.  For instance, try \verb sh ~ \verb ants.sh .
Additionally, if you run ants in tcsh shells, you will need to call it
as follows:
\begin{verbatim}
~/Code/ANTSSourceForgeSVN/bin/ANTS 2 -m PR\[r16slice.nii,r64slice.nii,1,2\]
\end{verbatim}
instead of
\begin{verbatim}
~/Code/ANTSSourceForgeSVN/bin/ANTS 2 -m PR[r16slice.nii,r64slice.nii,1,2]
\end{verbatim}
because tcsh does no interpret the brackets as intended.  Using the
bash shell script interface also avoids this.  The key change involves using the slash to force tcsh to 
interpet the bracket literally. 
\newpage
\section{ImageRegistration with ANTS}
\begin{wrapfigure}{r}{0.5\textwidth}
\framebox[0.5\textwidth]{
\begin{minipage}[t]{0.45\textwidth}
{\bf Quick Start:} {\color{blue}call ants.sh (from ANTS/Scripts/) to get usage and apply a normalization to some of your existing data.  It is instructional to read the script, modify some of the parameters and re-run to witness the effect of your changes -- image registration is an art as well as science.  Many other ``ready to go'' scripts are available in ANTS/Scripts. The user must set his/her ANTSPATH environment variable  -- which points to the location of ANTS binaries -- within these scripts or in their environment.  Note: All ANTS programs provide usage when called from the command line.  Most require the image dimension to be specified as the first parameter.  E.g. 
  ImageMath ImageDimension 
where ImageDimenion is 2 or 3.}
\end{minipage}}
\end{wrapfigure}
There are two general applications of image registration. The first application is transforming labeled data from a template image space into an individual space. This strategy is important when appearance alone is not enough to locate a structure as, for example, in the case of hippocampus segmentation. The template provides a prediction of the hippocampus-amygdala boundary.
The second application operates in the "inverse" direction of the first: instead of mapping template to individual, we map individual(s) to the template. Voxel-based population studies of either functional or structural variables depend on mapping to a template space. The common coordinate system enables a statistical evaluation of the likelihood of consistent activation across a group or, in other contexts, the differences in anatomy between two groups.

@@

\subsection{ANTS Transformation Models}
The ANTS toolkit provides a hierarchy of transformations with
adjustable levels of complexity, regularization, degrees of freedom
and behavior as optimizers.  The simplest transformation model is the
rigid and/or affine transform.  The most complex -- and most flexible
-- is a symmetric diffeomorphic transformation based on optimizing and
integrating a time-varying velocity field.  Computation time also
increases with transformation model complexity, but most normalization
needs may be met with under an hour of computation.  
We provide an overview of some of the ANTS models below and 
try to communicate intuition on what one gains/loses with each choice.  
An overview of available models and similarity terms is in Table~\ref{table:chart}.
\begin{table*}
  \centering
    \begin{tabular}{c c c c}
    {\bf Category} & {\bf Transformation, $\phi$} & {\bf Similarity Measures} & {\bf Brief Description} \\
    \toprule     
    \multirow{2}*{\bf Linear}
           & Rigid$^\dagger$ & MI, MSQ & Rigid registration. \\
       {} & Affine$^\dagger$ & MI, MSQ & Affine registration. \\
       \cmidrule(l){2-4}
    \multirow{2}*{\bf Elastic}
           & Deformable & CC, PR, MI, MSQ, PSE & Demons-like algorithm. \\
       {} & DMFFD & CC, PR, MI, MSQ, PSE  & FFD variant. \\
       \cmidrule(l){2-4}
    \multirow{2}*{\bf Diffeo.}
           & Exponential$^\dagger$ &  CC,PR, MI, MSQ, PSE  & $~\min~\velocity(\x)$ \\
       {} & Greedy SyN$^\dagger$ &  CC, PR, MI, MSQ, PSE  & ~~locally in time~$\min~\velocity(\x,t)$\\
       {} & Geodesic SyN$^\dagger$ &  CC, PR, MI, MSQ, PSE  &  $~\min~\velocity(\x,t)$ over all time \\
    \bottomrule
    \end{tabular}
  \caption{Transformations and a subset of the similarity metrics available in ANTS.  Similarity metric acronyms:  CC = fast cross correlation, PR = pure cross correlation (the preferred metric), MSQ = mean squared difference, MI = mutual information, PSE = point-set expectation \cite{Pluta2008}.  ANTS also provides the inverse of those transformations denoted by the `$\dagger$' symbol.  The brief descriptions of the diffeomorphic algorithms contrast the way in which the velocity field is optimized and used to parameterize $\phi$, the mapping.  All 
ANTS Diff algorithms generate $\phi(\x,t)$ over $t \in [ 0, 1]$ through gradient descent.}
  \label{table:chart}
\end{table*}    
\vspace{0in}

@@

\noindent{\bf Elastic/Vector Space Transformations.} If we assume no affine transformation, then 
an elastic transformation involves computing a mapping from image $I(\x)$ to image $J(\x)$ 
through a deformation field $\disp( \x )$.  
\begin{figure}
\center \includegraphics[width=0.75\textwidth]{./Figures/PullBack.pdf} 
\itkcaption[ANTS Deformation]{A deformation in a digital domain.}
\label{fig:pull}
\end{figure}  
The deformation is defined in the physical space of the image 
and dictates the positional difference between corresponding features in the two images.  
Thus, if a feature defined at $I(\x)$ matches a feature in $J$ at position $\y$ then the deformation 
field at $\x$ should give $\disp ( \x )=\y-\x$.   
Such a deformation field may be applied to deform 
image $J$ into image $I$ by composing the mapping $J_{deformed}(\x)=J(\x+\disp(\x))$.  
In a perfect world, then $I(\x)=J_{deformed}(\x)$, though this is rarely the case.  Figure~\ref{fig:pull} 
visualizes the deformation of an image under this standard model. 
Gradient descent optimization of an elastic mapping may be summarized (crudely) as: 
\begin{align}
\text{Compute the similarity gradient:   }  \nabla E = \partial_{\disp} \Pi(I,J(\x+\disp(\x))). \nonumber \\
\text{Update the deformation field:  }  \disp(\x) \leftarrow \disp(\x)+\delta \nabla E. \nonumber \\ 
\text{Regularize the deformation field:  }  \disp(\x) \leftarrow G_\sigma \star \disp(\x),
\end{align}
where $\Pi$ is the similarity, $\delta$ is a gradient step length and $G_\sigma$ is a gaussian smoother.
This optimization is captured in the following ANTS command:
\begin{verbatim}
ANTS 3 -m PR[AtlasHead.nii,crAnatomical.nii,1,4] -r Gauss[0,3]  
-o ElasticTest.nii.gz -i 30x20x10  -t Elast[1.5] 
\end{verbatim}
Here, we use the PR metric (a cross-correlation implementation) with
window radius 4, weight 1 and gradient step length 1.5.  The
optimization will be performed over three resolutions with a maximum
of 30 iterations at the coarsest level, 20 at the next coarsest and 10
at the full resolution.  We use a Gaussian regularizer with a sigma of
3 that operates only on the deformation field and not on the similarity gradient, as 
0 is passed as the first parameter to \verb Gauss .  One may see the
correspondence, yet again, between the ANTS call and the optimization
scheme.  The optimization will stop when either the energy cannot be 
smoothly minimized or the maximum number of iterations is reached. 
BSpline regularization is also available in ANTS -- see the DMFFD 
section below.

\noindent{\bf Warping and Invertibility.}  ANTS uses physical space to
define mappings. The origin etc is in the coordinates of the world in
which the picture (e.g. MRI) was taken. Then, warping between physical
coordinates is relatively easy. Differences in bounding boxes etc
present no problem -- assuming you avoid inconsistent headers
i.e. origins, directions, data orientation. One may use PrintHeader to
check the data and run simple, fast tests (few iterations) to perform
sanity checks before running through loads of data.
An ANTS deformation consists of
a standard naming prefix plus a standard naming extension. We usually 
assume nii but other file types may be used. The standard naming
extension is Warpxvec.nii for the x component of the deformation with
similar naming for y and z. Each component is stored in a separate
scalar image. The value of a voxel of a deformation/warp component
image is the physical space displacement from that voxel. Note that
the inverse mapping is stored in InverseWarpxvec.nii and provides the
mapping in the opposite direction.  Note that an inverse -- in the 
digital domain -- is only approximate as shown in figure~\ref{fig:inv}.
\begin{figure}
\center \includegraphics[width=0.75\textwidth]{Figures/DeformationsInDigitalDomain.pdf} 
\itkcaption[Digital Invertibility]{Digital invertibility presents some limitations. 
Here, we see that invertibility is not exact but is gained only by interpolation.  
Thus, in three-dimensional scenarios in particular, there are fundamental limits to the 
degree of invertibility that may be achieved.  The second and third voxels -- from left -- 
in image A undergo an expansion by a factor of 2.  That is, under the mapping, 
2 voxels are mapped to 4.  This gives the definition of the Jacobian -- computed 
by  ANTSJacobian -- which is a unitless measure defined by the 
ratio of volumes.  Thus, $\mathcal{J}(\x)=V(\phi(\x))/V(\x)$ where $V$ represents 
the volume operation and $\x$, here, may be a small object.  Thus, if $\phi$ -- the mapping -- 
causes expansion, then $\mathcal{J}(\x) > 1$. }
\vspace{-0.1in}
\label{fig:inv}
\end{figure}  
A few important notes follow. 
(1) Deformation directionality: Warps/deformations applied to images occur in the opposite direction from warps/deformations applied to points. That is, if we map image B to A using 
\begin{verbatim}
ANTS 2 -m PR[A,B,1,2] -o OUTPUT 
\end{verbatim}
 then we get a deformation called \verb OUTPUTWarp+extensions ~and an affine transform called \verb OUTPUTAffine.txt . 
This composite mapping - when applied to B - will transform B into the space of A. 
However, if we have points defined in B that we want to map to A, 
we have to use \verb OUTPUTInverseWarp ~and the inverse of \verb OUTPUTAffine.txt .  
This is because the domain and range of the map's definition need to be dense in a way that is appropriate for the data type. 
Figure~\ref{fig:inv} illustrates the concept. 
(2) Older Image Formats: older image formats (e.g. Analyze) may not
have proper origin/offset/direction support! In these cases, we
recommend converting to nii and verifying that data overlays properly.
(3) Transorm Composition: Composition of transforms may be achieved with
ComposeMultiTransform.  (4) Warping with inverses and concatenations --
viable when using diffeomorphisms -- are described in
figure~\ref{fig:warp}.
 \begin{figure}
\includegraphics[width=0.9\textwidth]{Figures/ANTSSyntax2.pdf} 
\itkcaption[ANTS Warping]{Here, we detail how one applies 
a deformation and associated inverses.  We also see how 
WarpImageMultiTransform can be used to concatenate a series 
of transformations.}
\vspace{-0.1in}
\label{fig:warp}
\end{figure}

@@

\subsection{ANTS Similarity Terms}
Here is a script that will let you experiment with different similarity term 
and transformation model combinations.  A few are listed, with parameters 
that are a reasonable starting point.
\begin{verbatim}
II=r16slice.nii # change to your own images 
JJ=r64slice.nii 
DIM=2 # two-dimensional images 
ITS=" -i 100x100x30 " # 3 optimization levels 
TSYNWITHTIME=" -t SyN[1,2,0.05]  -r Gauss[3,0.5] "
TGREEDYSYN=" -t SyN[0.25]  -r Gauss[3,0] "
TELAST=" -t Elast[1] -r Gauss[0,3] " 
TEXP="  -t Exp[0.5,10] -r Gauss[0,3] " 
LABELGUIDED=" -m PSE[${II},${JJ},${II},${JJ},0.75,0.1,25,0,10] "
INTMSQ=" -m MSQ[${II},${JJ},1,0.1] "
INTMI=" -m MI[${II},${JJ},1,32] "
INTPR=" -m PR[${II},${JJ},1,4] "
NAME=TEST
INT=$INTPR
TRAN=$TSYN
ANTS $DIM -o $NAME $ITS  $TRAN  $INT
INVW=" -i ${NAME}Affine.txt ${NAME}InverseWarp.nii "
FWDW=" ${NAME}Warp.nii ${NAME}Affine.txt "
WarpImageMultiTransform  2 ${II} a.nii -R ${JJ} $INVW
WarpImageMultiTransform  2 ${JJ} b.nii -R ${II} $FWDW
\end{verbatim}
Convergence occurs under two conditions: (1) either the maximum number of 
iterations are reached at a given optimization level or (2) the slope 
of the change in the optimization objective is negative or very small. 
{\em Note: ANTS does not like spaces between brackets in the call to the metrics.}

@@

A second type of multivariate normalization is able to use the T1 and DTI components 
directly in the optimization of the mapping.  An example of this type of call is below, 
where we assume that one has resampled the DT data to the space of T1: 
\begin{verbatim}
ANTS 3 -m PR[T1template.nii,T1subject.nii,1.25,4]
       -m PR[FAtemplate.nii,FAsubject.nii,1,4] 
-o MultiVar -i 30x30x20  -r Gauss[3,0] -t SyN[0.25]
\end{verbatim}
In this case, dense features within white matter are gained by using the DT component  
in addition to the T1 component.   The T1 PR metric, here, is weighted slightly more 
than the FA PR metric, 1.25 vs. 1.0.  {\em The convergence criterion for 
the multivariate scenario is a slave to the {\bf last} metric you 
pass on the ANTS command line}.

@@

\verb  sh ~ \verb buildtemplate.sh ~ to get usage. 
\item If you follow the usage, you will then call the script something like this: 
\verb sh ~ \verb buildtemplate.sh ~ \verb 2 ~ \verb B ~ \verb tiff ~ \verb OUT ~ \verb 4 ~ . 
\item The script should exit and ask if the parameters are ok for your problem.  
Check the path to the programs, the fact that all the images you want to use are being 
listed and that the template output name is ok with you.  
Once everything is prepared, then edit the script to remove the exit call at the end of the parameter check. 
\item  Call the script as in step 4 and then wait for the output.  The template will 
be called something like (in this example)  \verb OUTtemplate.nii .

@@

The script is fairly easy to alter so that you can send the whole thing to distributed computing. 
Typically, one would use voxbo or the qsub program available in the Sun Grid Engine.  This 
is what is done in \verb buildtemplateparallel.sh .  

@@

ANTS 3 -m PR[template.nii.gz,subject.nii.gz,1,2] -i 10x50x50x20
 -o subjectmap.nii -t SyN[0.25] -r Gauss[3,0]

@@

ANTS 3 -m PR[template.nii.gz,subject.nii.gz,1,4] -i 100x100x100x20
 -o subjectmap.nii -t SyN[0.25] -r Gauss[3,0] 

@@

A basic MRF segmentation algorithm is available in ANTS ImageMath.  We apply 

@@

ImageMath 2 r16slice.nii Segment r16slice.nii 4 0 
% output = r16slice_seg.nii, r16slice_prob_0.nii, r16slice_prob_1.nii, r16slice_prob_2.nii .  

@@

The parameter \verb 4 ~indicates 3 tissues plus background are sought.  The \verb 0 ~indicates 
that there are no prior images available to guide the segmentation.  Output is shown 

@@

ImageMath 2 r16sliceprior.nii Segment r16slice.nii 4 0.5 
   csfprior.nii wmprior.nii  gmprior.nii 
% output = r16sliceprior_seg.nii  r16sliceprior_prob_0.nii  
% r16sliceprior_prob_1.nii  r16sliceprior_prob_2.nii .  

@@

The parameter \verb 4 indicates 3 tissues plus background are sought.  The \verb 0.5 indicates 

>>>>

\subsection{The ANTS Executable}
The ANTS program itself is the central program encapsulating
normalization/registration functionality.  Its main output is an
affine transform file and a deformation field, potentially with
inverse.  Options to ANTS allows the user to navigate the similarity
and transformation options that are available.  ANTS takes multiple
similarity and optimization criteria as options.  The program is
wrapped in \verb ants.sh ~for normalization with ``out of the box''
parameters and in \verb buildtemplateparallel.sh ~for 
computationally distributed optimal template construction.
You can also call  \verb ANTS \verb --help to get basic usage.

\subsection{The WarpImageMultiTransform Executable}
The WarpImageMultiTransform program applies ANTS mappings to images.  One may apply an 
arbitrarily long series of transformations to an image through this program.  Thus, it enables one 
to compose a series of affine and deformable mappings -- and/or their inverses -- such that one 
may avoid multiple interpolations of a single image.  This program is used internally in ANTS 
optimization to avoid ever using more than one interpolation.  The use-NN option applies nearest 
neighbor interpolation, otherwise linear interpolation is used. 
You can call  \verb WarpImageMultiTransform \verb --help to get basic usage.

@@

\subsection{The ImageMath Executable} This is a multi-purpose program
that has the following syntax:  ``ImageMath ImageDimension
outputfilename   Operation  InputFileName  parameters ``.  Most basic
scalar image operations -- and some tensor operations -- may be
performed with this program.  Some operations output text files, some
output images and some output only to the terminal.   ImageMath allows
one to multiply images together (m), to negate images (Neg), to take
an image to a power (pow), to test the invertibility of
transformations (InvId), to compute the fractional anisotropy of an
image (TensorFA) and to perform Markov Random Field segmentation with
or without priors (Segment).   Many other operations are available.
Like all other ANTS programs, one may call ImageMath from the command
line to see all of its options.   ImageMath is used heavily in ANTS
scripts.   You can call it, in compination with grep or less, to attempt to
find some of its capabilities.

\subsection{ANTS Scripts} 
The ANTS/Scripts directory contains (hopefully) user-friendly
wrappings of ANTS tools that enable higher-level error checking and
combinations of basic ANTS functions.  These scripts are called as
\verb sh ~ \verb antsscriptname.sh ~ and provide usage when called
from the command line.  For instance, try \verb sh ~ \verb ants.sh .
Additionally, if you run ants in tcsh shells, you will need to call it
as follows:
\begin{verbatim}
~/Code/ANTSSourceForgeSVN/bin/ANTS 2 -m CC\[r16slice.nii,r64slice.nii,1,2\]
\end{verbatim}
instead of
\begin{verbatim}
~/Code/ANTSSourceForgeSVN/bin/ANTS 2 -m CC[r16slice.nii,r64slice.nii,1,2]
\end{verbatim}
because tcsh does no interpret the brackets as intended.  Using the
bash shell script interface also avoids this.  The key change involves using the slash to force tcsh to 
interpet the bracket literally. 
\newpage
\section{ImageRegistration with ANTS}
\begin{wrapfigure}{r}{0.5\textwidth}
\framebox[0.5\textwidth]{
\begin{minipage}[t]{0.45\textwidth}
{\bf Quick Start:} {\color{blue}call ants.sh (from ANTS/Scripts/) to get usage and apply a normalization to some of your existing data.  It is instructional to read the script, modify some of the parameters and re-run to witness the effect of your changes -- image registration is an art as well as science.  Many other ``ready to go'' scripts are available in ANTS/Scripts. The user must set his/her ANTSPATH environment variable  -- which points to the location of ANTS binaries -- within these scripts or in their environment.  Note: All ANTS programs provide usage when called from the command line.  Most require the image dimension to be specified as the first parameter.  E.g. 
  ImageMath ImageDimension 
where ImageDimenion is 2 or 3.}
\end{minipage}}
\end{wrapfigure}
There are two general applications of image registration. The first application is transforming labeled data from a template image space into an individual space. This strategy is important when appearance alone is not enough to locate a structure as, for example, in the case of hippocampus segmentation. The template provides a prediction of the hippocampus-amygdala boundary.
The second application operates in the "inverse" direction of the first: instead of mapping template to individual, we map individual(s) to the template. Voxel-based population studies of either functional or structural variables depend on mapping to a template space. The common coordinate system enables a statistical evaluation of the likelihood of consistent activation across a group or, in other contexts, the differences in anatomy between two groups.

@@

\subsection{ANTS Transformation Models}
The ANTS toolkit provides a hierarchy of transformations with
adjustable levels of complexity, regularization, degrees of freedom
and behavior as optimizers.  The simplest transformation model is the
rigid and/or affine transform.  The most complex -- and most flexible
-- is a symmetric diffeomorphic transformation based on optimizing and
integrating a time-varying velocity field.  Computation time also
increases with transformation model complexity, but most normalization
needs may be met with under an hour of computation.  
We provide an overview of some of the ANTS models below and 
try to communicate intuition on what one gains/loses with each choice.  
An overview of available models and similarity terms is in Table~\ref{table:chart}.
\begin{table*}
  \centering
    \begin{tabular}{c c c c}
    {\bf Category} & {\bf Transformation, $\phi$} & {\bf Similarity Measures} & {\bf Brief Description} \\
    \toprule     
    \multirow{2}*{\bf Linear}
           & Rigid$^\dagger$ & MI, MSQ & Rigid registration. \\
       {} & Affine$^\dagger$ & MI, MSQ & Affine registration. \\
       \cmidrule(l){2-4}
    \multirow{2}*{\bf Elastic}
           & Deformable & CC, PR, MI, MSQ, PSE & Demons-like algorithm. \\
       {} & DMFFD & CC, PR, MI, MSQ, PSE  & FFD variant. \\
       \cmidrule(l){2-4}
    \multirow{2}*{\bf Diffeo.}
           & Exponential$^\dagger$ &  CC,PR, MI, MSQ, PSE  & $~\min~\velocity(\x)$ \\
       {} & Greedy SyN$^\dagger$ &  CC, PR, MI, MSQ, PSE  & ~~locally in time~$\min~\velocity(\x,t)$\\
       {} & Geodesic SyN$^\dagger$ &  CC, PR, MI, MSQ, PSE  &  $~\min~\velocity(\x,t)$ over all time \\
    \bottomrule
    \end{tabular}
  \caption{Transformations and a subset of the similarity metrics
    available in ANTS.  Similarity metric acronyms:  CC = fast cross
    correlation (the preferred metric), PR = cross correlation (deprecated), MSQ = mean squared difference, MI = mutual information, PSE = point-set expectation \cite{Pluta2008}.  ANTS also provides the inverse of those transformations denoted by the `$\dagger$' symbol.  The brief descriptions of the diffeomorphic algorithms contrast the way in which the velocity field is optimized and used to parameterize $\phi$, the mapping.  All 
ANTS Diff algorithms generate $\phi(\x,t)$ over $t \in [ 0, 1]$ through gradient descent.}
  \label{table:chart}
\end{table*}    
\vspace{0in}

@@

\noindent{\bf Elastic/Vector Space Transformations.} If we assume no affine transformation, then 
an elastic transformation involves computing a mapping from image $I(\x)$ to image $J(\x)$ 
through a deformation field $\disp( \x )$.  
\begin{figure}
\center \includegraphics[width=0.75\textwidth]{./Figures/PullBack.pdf} 
\itkcaption[ANTS Deformation]{A deformation in a digital domain.}
\label{fig:pull}
\end{figure}  
The deformation is defined in the physical space of the image 
and dictates the positional difference between corresponding features in the two images.  
Thus, if a feature defined at $I(\x)$ matches a feature in $J$ at position $\y$ then the deformation 
field at $\x$ should give $\disp ( \x )=\y-\x$.   
Such a deformation field may be applied to deform 
image $J$ into image $I$ by composing the mapping $J_{deformed}(\x)=J(\x+\disp(\x))$.  
In a perfect world, then $I(\x)=J_{deformed}(\x)$, though this is rarely the case.  Figure~\ref{fig:pull} 
visualizes the deformation of an image under this standard model. 
Gradient descent optimization of an elastic mapping may be summarized (crudely) as: 
\begin{align}
\text{Compute the similarity gradient:   }  \nabla E = \partial_{\disp} \Pi(I,J(\x+\disp(\x))). \nonumber \\
\text{Update the deformation field:  }  \disp(\x) \leftarrow \disp(\x)+\delta \nabla E. \nonumber \\ 
\text{Regularize the deformation field:  }  \disp(\x) \leftarrow G_\sigma \star \disp(\x),
\end{align}
where $\Pi$ is the similarity, $\delta$ is a gradient step length and $G_\sigma$ is a gaussian smoother.
This optimization is captured in the following ANTS command:
\begin{verbatim}
ANTS 3 -m CC[AtlasHead.nii,crAnatomical.nii,1,4] -r Gauss[0,3]  
-o ElasticTest.nii.gz -i 30x20x10  -t Elast[1.5] 
\end{verbatim}
Here, we use the CC metric (a cross-correlation implementation) with
window radius 4, weight 1 and gradient step length 1.5.  The
optimization will be performed over three resolutions with a maximum
of 30 iterations at the coarsest level, 20 at the next coarsest and 10
at the full resolution.  We use a Gaussian regularizer with a sigma of
3 that operates only on the deformation field and not on the similarity gradient, as 
0 is passed as the first parameter to \verb Gauss .  One may see the
correspondence, yet again, between the ANTS call and the optimization
scheme.  The optimization will stop when either the energy cannot be 
smoothly minimized or the maximum number of iterations is reached. 
BSpline regularization is also available in ANTS -- see the DMFFD 
section below.

\noindent{\bf Warping and Invertibility.}  ANTS uses physical space to
define mappings. The origin etc is in the coordinates of the world in
which the picture (e.g. MRI) was taken. Then, warping between physical
coordinates is relatively easy. Differences in bounding boxes etc
present no problem -- assuming you avoid inconsistent headers
i.e. origins, directions, data orientation. One may use PrintHeader to
check the data and run simple, fast tests (few iterations) to perform
sanity checks before running through loads of data.
An ANTS deformation consists of
a standard naming prefix plus a standard naming extension. We usually 
assume nii but other file types may be used. The standard naming
extension is Warpxvec.nii.gz for the x component of the deformation with
similar naming for y and z. Each component is stored in a separate
scalar image. The value of a voxel of a deformation/warp component
image is the physical space displacement from that voxel. Note that
the inverse mapping is stored in InverseWarpxvec.nii.gz and provides the
mapping in the opposite direction.  Note that an inverse -- in the 
digital domain -- is only approximate as shown in figure~\ref{fig:inv}.
\begin{figure}
\center \includegraphics[width=0.75\textwidth]{Figures/DeformationsInDigitalDomain.pdf} 
\itkcaption[Digital Invertibility]{Digital invertibility presents some limitations. 
Here, we see that invertibility is not exact but is gained only by interpolation.  
Thus, in three-dimensional scenarios in particular, there are fundamental limits to the 
degree of invertibility that may be achieved.  The second and third voxels -- from left -- 
in image A undergo an expansion by a factor of 2.  That is, under the mapping, 
2 voxels are mapped to 4.  This gives the definition of the Jacobian -- computed 
by  ANTSJacobian -- which is a unitless measure defined by the 
ratio of volumes.  Thus, $\mathcal{J}(\x)=V(\phi(\x))/V(\x)$ where $V$ represents 
the volume operation and $\x$, here, may be a small object.  Thus, if $\phi$ -- the mapping -- 
causes expansion, then $\mathcal{J}(\x) > 1$. }
\vspace{-0.1in}
\label{fig:inv}
\end{figure}  
A few important notes follow. 
(1) Deformation directionality: Warps/deformations applied to images occur in the opposite direction from warps/deformations applied to points. That is, if we map image B to A using 
\begin{verbatim}
ANTS 2 -m CC[A,B,1,2] -o OUTPUT 
\end{verbatim}
 then we get a deformation called \verb OUTPUTWarp+extensions ~and an affine transform called \verb OUTPUTAffine.txt . 
This composite mapping - when applied to B - will transform B into the space of A. 
However, if we have points defined in B that we want to map to A, 
we have to use \verb OUTPUTInverseWarp ~and the inverse of \verb OUTPUTAffine.txt .  
This is because the domain and range of the map's definition need to be dense in a way that is appropriate for the data type. 
Figure~\ref{fig:inv} illustrates the concept. 
(2) Older Image Formats: older image formats (e.g. Analyze) may not
have proper origin/offset/direction support! In these cases, we
recommend converting to nii and verifying that data overlays properly.
(3) Transorm Composition: Composition of transforms may be achieved with
ComposeMultiTransform.  (4) Warping with inverses and concatenations --
viable when using diffeomorphisms -- are described in
figure~\ref{fig:warp}.
 \begin{figure}
\includegraphics[width=0.9\textwidth]{Figures/ANTSSyntax2.pdf} 
\itkcaption[ANTS Warping]{Here, we detail how one applies 
a deformation and associated inverses.  We also see how 
WarpImageMultiTransform can be used to concatenate a series 
of transformations.}
\vspace{-0.1in}
\label{fig:warp}
\end{figure}

@@

\subsection{ANTS Similarity Terms}
Here is a script that will let you experiment with different similarity term 
and transformation model combinations.  A few are listed, with parameters 
that are a reasonable starting point.
\begin{verbatim}
II=r16slice.nii # change to your own images 
JJ=r64slice.nii 
DIM=2 # two-dimensional images 
ITS=" -i 100x100x30 " # 3 optimization levels 
TSYNWITHTIME=" -t SyN[0.5,5,0.01]  -r Gauss[3,0] "
TGREEDYSYN=" -t SyN[0.25]  -r Gauss[3,0] "
TELAST=" -t Elast[1] -r Gauss[0.5,3] " 
TEXP="  -t Exp[0.5,10] -r Gauss[0.5,3] " 
LABELGUIDED=" -m PSE[${II},${JJ},${II},${JJ},0.75,0.1,25,0,10] "
INTMSQ=" -m MSQ[${II},${JJ},1,0] "
INTMI=" -m MI[${II},${JJ},1,32] "
INTCC=" -m CC[${II},${JJ},1,4] "
NAME=TEST
INT=$INTPR
TRAN=$TSYN
ANTS $DIM -o $NAME $ITS  $TRAN  $INT
INVW=" -i ${NAME}Affine.txt ${NAME}InverseWarp.nii.gz "
FWDW=" ${NAME}Warp.nii.gz ${NAME}Affine.txt "
WarpImageMultiTransform  2 ${II} a.nii.gz -R ${JJ} $INVW
WarpImageMultiTransform  2 ${JJ} b.nii.gz -R ${II} $FWDW
\end{verbatim}
Convergence occurs under two conditions: (1) either the maximum number of 
iterations are reached at a given optimization level or (2) the slope 
of the change in the optimization objective is negative or very small. 
{\em Note: ANTS does not like spaces between brackets in the call to the metrics.}

@@

A second type of multivariate normalization is able to use the T1 and DTI components 
directly in the optimization of the mapping.  An example of this type of call is below, 
where we assume that one has resampled the DT data to the space of T1: 
\begin{verbatim}
ANTS 3 -m CC[T1template.nii,T1subject.nii,1.25,4]
       -m CC[FAtemplate.nii,FAsubject.nii,1,4] 
-o MultiVar -i 30x30x20  -r Gauss[3,0] -t SyN[0.25]
\end{verbatim}
In this case, dense features within white matter are gained by using the DT component  
in addition to the T1 component.   The T1 CC metric, here, is weighted slightly more 
than the FA CC metric, 1.25 vs. 1.0.  {\em The convergence criterion for 
the multivariate scenario is a slave to the {\bf last} metric you 
pass on the ANTS command line}.

@@

\verb  bash ~ \verb buildtemplateparallel.sh -h ~ to get usage. 
\item Following the usage, you can write a script that contains a call like this: 
\verb buildtemplateparallel.sh ~ \verb  -d ~ \verb  2 ~ \verb -o ~ \verb  BTP ~ \verb -c ~ \verb  0 ~ \verb  B*tiff
% \verb bash ~ \verb ~ \verb 2 ~ \verb B ~ \verb tiff ~ \verb OUT ~ \verb 4 ~ . 
\item  Call the script in step 4 and then wait for the output.  The template will 
be called (in this example)  \verb BTPtemplate.nii.gz .
\item If the script does not produce the desired output, check that
  you defined \$ANTSPATH in your path.
You may also check and or define the path to the ANTs programs
explicitly by editing this variable in buildtemplateparallel and antsIntroduction.

@@

The script offers an option ( -c 1  for qsub , -c 2 for voxbo )  so that you can send the whole thing to distributed computing. 
Typically, one would use voxbo or the qsub program available in the Sun Grid Engine.  

@@

ANTS 3 -m CC[template.nii.gz,subject.nii.gz,1,2] -i 10x50x50x20
 -o subjectmap.nii.gz -t SyN[0.25] -r Gauss[3,0]

@@

ANTS 3 -m CC[template.nii.gz,subject.nii.gz,1,4] -i 100x100x100x20
 -o subjectmap.nii.gz -t SyN[0.25] -r Gauss[3,0] 

@@

An advanced MRF segmentation algorithm is available in ANTS via the
\verb Atropos tool, which is still under development.  We apply 

@@

  ThresholdImage 2 r16slice.nii mask.nii.gz 0.001 Inf
  Atropos -d 2 -a r16slice.nii -o [r16slice_seg.nii.gz,r16slice_prob_%02d.nii.gz] \
      -c [10,0.001] -m [0.05,1] -i kmeans[3] -x mask.nii.gz

@@

The parameter \verb kmeans[3] ~indicates 3 tissues plus background are sought.  The \verb [0.05,1] ~indicates 
that some MRF smoothing will be applied to the labels.  Other
options are defined in the Atropos help.  Output is shown 

@@

  Atropos -d 2 -a r16slice.nii -o [r16slice_seg.nii.gz,r16slice_prob_%02d.nii.gz] \
      -c [10,0.001] -m [0.05,1] \ 
      -i  PriorProbabilityImages[3,r16slice_prob_%02d.nii.gz,0.5] -x mask.nii.gz

@@

The \verb PriorProbabilityImages ~parameter~ \verb 3 ~indicates 3 tissues
plus background are sought.  The \verb PriorProbabilityImages ~term \verb 0.5 ~indicates 
